Containers and R---the Rockerverse and beyond

### 2020-01-15 manuscript.tex
update .tex file

\abstract{%
The Rocker project provides widely-used Docker images for R across
different application scenarios. This articles surveys describe
downstream projects building upon Rocker. We also look beyond Rocker to
other projects connecting containerisation with R. These use cases and
the diversity of applications demonstrate the power of Rocker and
containerisation for collaboration, effectivity, scalability, and
transparency.
}

@@

The R community keeps growing. This can be seen in the number of new
packages on CRAN, which keeps on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies taking up,
as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. Coinciding
with the rise of R was the advent of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distribution and deployment of server applications.
Combining both these topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides images with R (see the
next Section for more details). The considerable uptake and continued
evolution of the Rocker Project has lead to numerous projects extending
or building upon Rocker images, ranging from reproducible research to
production deployments. This article presents this \emph{Rockerverse} of
projects across all development stages: early demonstrations, working
prototypes, and mature products. We also introduce related activities
connecting the R language and environment with other containerisation
solutions. The main contributions is a coherent picture of the current
lay of the land of using containers in, with, and for R.

@@

@@

The key accomplishment of Docker as an ``application'' is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive proposition and Docker's very easy to use user
interface has lead to widespread adoption and use of Docker in a variety
of domains, e.g., cloud computing infrastructure \citep{Bernstein2014},
data science \citep{boettiger_introduction_2015}, and edge computing
\citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
a natural match for ``cloud deployment'' which runs, or at least appears
to run, ``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
build from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

For statistical computing and analysis centered around R, the Rocker
Project has provided a variety of Docker containers since its start in
2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
containers spanning to from building blocks with \texttt{R-release} or
\texttt{R-devel}, via containers with RStudio\textasciitilde{}Server and
Shiny\textasciitilde{}Server, to domain-specific containers such as
\texttt{geospatial}. These containers form \emph{image stacks}, building
on top of each other for better maintainability and Also of note is a
series of ``versioned'' containers which match the R release they
contain with the \emph{then-current} set of packages via the MRAN
Snapshot views of CRAN \citep{microsoft_cran_2019}. The Rocker Project's
impact and importance was acknowledged by the Chan Zuckerberg
Initiative's \emph{Essential Open Source Software for Science}, who
provide findung for the projects's sustainable maintenance, community
growth, and targeting new hardware platforms including GPUs
\citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software. An alternative
stemming from the domain of high-perfomance computing is
\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
run Docker images, and in the case of Rocker works out of the box if the
main process is R, e.g., in \texttt{rocker/r-base}, but does not succeed
in running images where there is an init script, e.g., in containers
that by default run RStudio. In the latter case, a \texttt{Singularity}
file, a recipe akin to a \texttt{Dockerfile}, needs to be used. To date,
no comparable image stack to Rocker exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which also
can build \texttt{Dockerfile}s and run Docker images. Proof of concepts
for using podman to build and run Rocker containers
exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
Yet the prevelance of Docker, especially in the broader user community
beyond experts or niche systems, and the vast amount of blog posts and
courses for Docker, currently caps specific development efforts for both
Singularity and podman in the R community. This might quickly change
when usability and spread increase, or security features such as
rootless/unprivileged containers, which both these tools support out of
the box, become more sought after.

@@

@@

@@

@@

@@

@@

@@

\textbf{RStudio} also maintains a set of
\href{https://github.com/rstudio/r-docker}{Docker images} for multiple
operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
stacks include an \href{https://github.com/rstudio/r-builds}{opinionated
R installation} and ensure the R installation and profile is consistent
across the different Linux distributions. The images also install R in a
versioned directory and use a minimal set of build and runtime
dependencies, Furthermore these images are responsible for creating
pre-compiled binary R packages for Linux, which supplement the binaries
built by CRAN. These pre-compiled packages dramatically decrease the
installation time of R packages on Linux and subsequently can decrease
the build time of Docker images \citep{lopp_package_2019}.

@@

@@

@@

@@

\textbf{\pkg{stevedore}}
(\url{https://cran.r-project.org/package=stevedore})

@@

\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}

@@

@@

\begin{itemize}
\tightlist
\item
  \href{https://github.com/ColinFay/dockerfiler/}{\texttt{dockerfiler}}
\item
  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
\item
  \href{https://github.com/smaakage85/dockr}{\texttt{dockr}}, see also
  \url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}
\item
  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
\item
  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
  with Docker}
\item
  \url{https://github.com/gigantum/base-images}
\end{itemize}

@@

\hypertarget{deployment-to-cloud-services}{%
\subsection{Deployment to cloud
services}\label{deployment-to-cloud-services}}

The cloud is the natural environment of containers, and becomes the
go-to mechanism to expose applications written in R to users.

@@

@@

@@

@@

  \begin{itemize}
  \tightlist
  \item
    \url{https://cloud.google.com/cloud-build/}
  \item
    Cloud Build is another continuous intergation service that helps
    move the workload of building the Docker images to an online
    service. Cloud Build can be set up to build the Dockerfiles on each
    GitHub commit or release. This means you do not need to build the
    Docker images locally, which can tie up resources since Docker
    images can be several GBs and take a long time to compile. Google
    Cloud Build works alongside Google Container Registry to allow you
    to build private and public Docker images, which allows you to build
    up your own dependency graphs for downstream applications.
  \end{itemize}
\end{itemize}

@@

@@

Using \textbf{GPUs} (graphical processing units) as a specialised
hardware from containerized common work environments is also possible
and useful \citep{haydel_enhancing_2015}. GPUs are increasingly popular
for compute-intensive machine learning tasks, e.g., deep artificial
neural networks \citep{schmidhuber_deep_2015}. Though in this case,
containers are not completely portable between hardware environments.
Containers running GPU software require drivers and libraries specific
to GPU models and versions, and containers require a specialized runtime
to connect to the underlying GPU hardware. For NVIDIA GPUs, the
\href{https://github.com/NVIDIA/nvidia-docker}{NVIDIA Container Toolkit}
includes a specialized runtime plugin for Docker and a set of base
images with appropriate drivers and libraries. The Rocker project has
(beta) images based on these that include GPU-enabled versions of
machine-learning R packages: \url{https://github.com/rocker-org/ml}.

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{To deploy and manage RStudio Professional Products:} Many
  RStudio customers are managing professional products in Docker
  containers as an alternative to bare-metals servers or virtual
  machines. Best practices for
  \href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
  RStudio with Docker containers} as well as
  \href{https://github.com/rstudio/rstudio-docker-products}{Docker
  images} for RStudio's commerical products are available.
\item
  \emph{To run interactive RStudio sessions on a cluster:} In the
  RStudio Server Pro 1.2 release in 2019, RStudio added new
  functionality called
  \href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
  gives users the ability to spawn RStudio instances and background jobs
  in a scalable way on external clusters, e.g.,
  \href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
  based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
  clusters, and optionally, with Singularity containers. A key benefit
  of Launcher is the ability for R and Python users to interactively
  work in RStudio without learning about Docker at all - while still
  leveraging containers and Kubernetes. Users can submit batch jobs to
  Kubernetes clusters without writing specific deployment scripts.
\end{enumerate}

@@

@@

  \begin{itemize}
  \tightlist
  \item
    A popular application of using Docker is to create the environments
    that can run in parallel for speeding up R code. For example,
    \href{https://CRAN.R-project.org/package=googleComputeEngineR}{\texttt{googleComputeEngineR}}'s
    \texttt{gce\_vm\_cluster()} function can create clusters of 2 or
    more virtual machines, running multi-CPU architectures. Instead of
    running a local R script with the local CPU and RAM restrictions,
    the same code can be processed on all CPU threads of the cluster of
    machines in the cloud, all running in a Docker container with the
    same R environments. This is achieved through
    \texttt{googleComputeEngineR}'s integration with the R parralisation
    library
    \href{https://CRAN.R-project.org/package=future}{\texttt{library(future)}
    by Henrik Bengtsson}. Local R computation can be thrown up to a
    multi-CPU and VM environment to achieve parrell computation in a few
    lines of R code in your local session. -
    \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
    demonstrations are available here}.
  \end{itemize}
\item
  \href{https://cloud.run}{Google Cloud Run} - This is a CaaS
  (Containers as a Service) that lets you launch a Docker container
  without worrying about underlying infrastructure. This dispenses with
  the developer creating a cloud server to run the Docker image on, by
  abstracting away those servers to a more serverless configuration.
  Cloud Run lets you run your code on top of a managed or your own
  Kubernetes cluster running Knative, and can accept any Docker image.
  The service takes care of network ingress, scaling machines up and
  down to zero, authentication and authorisation, all features which are
  non-trivial for a developer to create on their own. This can be used
  to scale up R code to millions of instances if they need to, with
  little or no changes to existing code. An R implementation is shown
  here at
  \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
  uses Cloud Run to create a scalable R plumber API.
\item
  \texttt{batchtools} \citep{Lang2017batchtools} can
  \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
  jobs with Docker Swarm}
\item
  scalable deployments, e.g., start with numerous Shiny talks mentioning
  Rocker at useR!2017
\item
  \href{https://github.com/dynverse/dynmethods}{dynmethods}
  \citep{rcannood}: In order to evaluate ±50 computational methods which
  all used different environments (R, Python, C++, \ldots{}), we wrapped
  each of them in a docker container and can execute these methods from
  R. Again, all of these containers are being built on travis-ci, and
  will only be pushed to docker hub if the integration test succeeds.
\item
  \pkg{drake},
  \url{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}
\end{itemize}

\hypertarget{research-compendia-benmarwick}{%
\subsection{\texorpdfstring{Research Compendia
\citep{benmarwick}}{Research Compendia {[}@benmarwick{]}}}\label{research-compendia-benmarwick}}

\begin{itemize}
\tightlist
\item
  \url{https://github.com/benmarwick/rrtools}
\item
  \pkg{renv}, \url{https://rstudio.github.io/renv/articles/docker.html}
\end{itemize}

@@

Containers can also serve as useful playgrounds to create environments
ad-hoc or to provide very specific environments that are not often
needed. Developers can readily start a local container based on a Rocker
image to investigate a bug report, which might require them to install a
specific combination of package versions or use a specific R version
(see also versioned images above). Unlike recreating the reporter's
environment on their regular machine, a container can simply be
discarded after use, and there is no need to tediously roll back
explorative changes to a previous state. Using the Rocker images with
RStudio, the disposable environments lack no development comfort (see
also the usage in research compendia).

\textbf{\href{https://github.com/wch/r-debug}{\texttt{r-debug}}} is a
purpose built Docker image for debugging R memory problems.
\citet{eddelbuettel_debugging_2019} describes how a Docker container was
used to debug an issue with a package only occuring with a particular
version of Fortran, and using tools which are not readily available on
all platforms (e.g., not on macOS).

The R-hub Docker images (see Section~\nameref{rhub}) can also be used to
debug problems locally using various combinations of Linux platforms, R
versions, and compilers\textbackslash{}footnote\{See
\href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html}
and
\href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}.
The images go beyond the configurations, or \emph{flavours}, used by
CRAN for checking
packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}},
e.g., with CentOS-based images, but lack a container for checking on
Windows or OS X. The images greatly support package developers to
provide support on operating systems they are not familiar.

>>>>

\abstract{%
The Rocker project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. We also
look beyond Rocker to other projects connecting containerisation with R,
namely alternative suites of images. These use cases and the variety of
applications demonstrate the power of Rocker specifically and
containerisation in general. We identified common themes across this
diversity: reproducible environments, scalability and effectivity, and
portability across clouds.
}

@@

\label{intro}

The R community keeps growing. This can be seen in the number of new
packages on CRAN, which keeps on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies taking up,
as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. Coinciding
with the rise of R was the advent of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distribution and deployment of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides images with R (see the
next Section for more details). The considerable uptake and continued
evolution of the Rocker Project has lead to numerous projects extending
or building upon Rocker images, ranging from reproducible research to
production deployments. This article presents this \emph{Rockerverse} of
projects across all development stages: early demonstrations, working
prototypes, and mature products. We also introduce related activities
connecting the R language and environment with other containerisation
solutions. The main contributions is a coherent picture of the current
lay of the land of using containers in, with, and for R.

@@

\label{containerisation} \label{rocker}

@@

The key accomplishment of Docker as an ``application'' is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive proposition and Docker's very easy to use user
interface has lead to widespread adoption and use of Docker in a variety
of domains, e.g., cloud computing infrastructure
\citep[e.g.,][]{Bernstein2014}, data science
\citep[e.g.,][]{boettiger_introduction_2015}, and edge computing
\citep[e.g.,][]{alam_orchestration_2018}. It provided to be a natural
match for ``cloud deployment'' which runs, or at least appears to run,
``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
build from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

For statistical computing and analysis centered around R, the Rocker
Project has provided a variety of Docker containers since its start in
2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
containers spanning to from building blocks with \texttt{R-release} or
\texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability
(i.e.~smaller \texttt{Dockerfile}s), composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software. An alternative
stemming from the domain of high-perfomance computing is
\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
run Docker images, and in the case of Rocker works out of the box if the
main process is R, e.g., in \texttt{rocker/r-base}, but does not succeed
in running images where there is an init script, e.g., in containers
that by default run RStudio~Server. In the latter case, a
\texttt{Singularity} file, a recipe akin to a \texttt{Dockerfile}, needs
to be used. To date, no comparable image stack to Rocker exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which also
can build \texttt{Dockerfile}s and run Docker images. Proof of concepts
for using podman to build and run Rocker containers
exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
Yet the prevelance of Docker, especially in the broader user community
beyond experts or niche systems, and the vast amount of blog posts and
courses for Docker, currently caps specific development efforts for both
Singularity and podman in the R community. This might quickly change
when usability and spread increase, or security features such as
rootless/unprivileged containers, which both these tools support out of
the box, become more sought after.

@@

\label{images}

@@

\label{alternatives}

@@

\label{bioc}

@@

\label{versions}

@@

\label{windows}

@@

\label{nondebian}

@@

\href{https://rstudio.com/}{\textbf{RStudio}} also maintains a set of
\href{https://github.com/rstudio/r-docker}{Docker images} for multiple
operating systems including SUSE, Ubuntu, CentOS, and Debian. The image
stack includes an \href{https://github.com/rstudio/r-builds}{opinionated
R installation} and ensures the R installation and profile is consistent
across the different Linux distributions. The base image also installs R
in a versioned directory and uses a minimal set of build and runtime
dependencies. Furthermore these images are responsible for creating
pre-compiled binary R packages for Linux, which supplement the binaries
built by CRAN. These pre-compiled packages dramatically decrease the
installation time of R packages on Linux and subsequently can decrease
the build time of Docker images \citep{lopp_package_2019}.

@@

\label{datascience}

@@

\textbf{\href{http://gigantum.com/}{Gigantum}} also its own image stack
(\url{https://github.com/gigantum/base-images}). \ldots{}

@@

\label{applications}

@@

\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor} is not
actively maintained anymore, but should be honorably mentioned as the
first R package for managing Docker images and containers. It uses the
\pkg{sys} package to run system commands against the Docker CLI, both
locally and through an ssh connection, and has convenience functions,
e.g., for listing and removing containers/images and for accessing logs.
The output of container executions are converted to appropriate R types.
The Docker CLI, while evolving quickly and not with great concern for
introducing breaking changes, basic functionality is unchanged for a
long time so \texttt{harbor::docker\_run(image\ =\ "hello-world")} works
like a charm.

\textbf{\pkg{stevedore}}
(\url{https://cran.r-project.org/package=stevedore}) currently the most
powerful Docker client in R.

@@

@@

\label{envs}

@@

\textbf{\pkg{dockerfiler}}
(\url{https://github.com/ColinFay/dockerfiler/}) \ldots{}

\textbf{\pkg{containerit}}
(\url{https://github.com/o2r-project/containerit/}) \ldots{}

\textbf{\pkg{dockr}} (\url{https://github.com/smaakage85/dockr})
\ldots{} see/cite
\url{http://smaakage85.netlify.com/2019/12/21/dockr-easy-containerization-for-r/}

\begin{itemize}
\tightlist
\item
  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
  with Docker}
\end{itemize}

@@

@@

\label{enterprise}

@@

\label{deployment}

The cloud is the natural environment of containers, and is the natural
mechanism to deploy R server applications.

@@

\label{cicd}

@@

  \begin{itemize}
  \tightlist
  \item
    \url{https://cloud.google.com/cloud-build/}
  \item
    Cloud Build is another continuous intergation service that helps
    move the workload of building the Docker images to an online
    service. Cloud Build can be set up to build the Dockerfiles on each
    GitHub commit or release. This means you do not need to build the
    Docker images locally, which can tie up resources since Docker
    images can be several GBs and take a long time to compile. Google
    Cloud Build works alongside Google Container Registry to allow you
    to build private and public Docker images, which allows you to build
    up your own dependency graphs for downstream applications.
  \item
    \url{https://code.markedmondson.me/googleCloudRunner/articles/cloudbuild.html}
  \end{itemize}
\end{itemize}

@@

\label{workenvs}

@@

Using \textbf{GPUs} (graphical processing units) as a specialised
hardware from containerized common work environments is also possible
and useful \citep{haydel_enhancing_2015}. GPUs are increasingly popular
for compute-intensive machine learning (ML) tasks, e.g., deep artificial
neural networks \citep{schmidhuber_deep_2015}. Though in this case,
containers are not completely portable between hardware environments,
but the software stack for ML with GPUs is so complex to set up that a
ready-to-use container is helpful. Containers running GPU software
require drivers and libraries specific to GPU models and versions, and
containers require a specialized runtime to connect to the underlying
GPU hardware. For NVIDIA GPUs, the
\href{https://github.com/NVIDIA/nvidia-docker}{NVIDIA Container Toolkit}
includes a specialized runtime plugin for Docker and a set of base
images with appropriate drivers and libraries. The Rocker project
\href{https://github.com/rocker-org/ml}{has a repository} with (beta)
images based on these that include GPU-enabled versions of
machine-learning R packages, e.g., \texttt{rocker/ml} and
\texttt{rocker/tensorflow-gpu}.

@@

\label{rstudio}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \emph{To deploy and manage RStudio Professional Products:} Many
  RStudio customers are managing professional products in Docker
  containers as an alternative to bare-metals servers or virtual
  machines. Best practices for
  \href{https://support.rstudio.com/hc/en-us/articles/360021594513-Running-RStudio-with-Docker-containers}{running
  RStudio with Docker containers} as well as
  \href{https://github.com/rstudio/rstudio-docker-products}{Docker
  images} for RStudio's commerical products are available.
\item
  \emph{To run interactive RStudio sessions on a cluster:} In the
  RStudio Server Pro 1.2 release in 2019, RStudio added new
  functionality called
  \href{https://solutions.rstudio.com/launcher/overview/}{Launcher}. It
  gives users the ability to spawn R sessions and background jobs in a
  scalable way on external clusters, e.g.,
  \href{https://support.rstudio.com/hc/en-us/articles/360019253393-Using-Docker-images-with-RStudio-Server-Pro-Launcher-and-Kubernetes}{Kubernetes
  based on Docker images} or \href{https://slurm.schedmd.com/}{Slurm}
  clusters, and optionally, with Singularity containers. A key benefit
  of Launcher is the ability for R and Python users to interactively
  work in RStudio without learning about Docker at all - while still
  leveraging containers and Kubernetes. Users can submit batch jobs to
  Kubernetes clusters without writing specific deployment scripts.
\end{enumerate}

@@

\label{processing}

@@

  \begin{itemize}
  \tightlist
  \item
    A popular application of using Docker is to create the environments
    that can run in parallel for speeding up R code. For example,
    \href{https://CRAN.R-project.org/package=googleComputeEngineR}{\texttt{googleComputeEngineR}}'s
    \texttt{gce\_vm\_cluster()} function can create clusters of 2 or
    more virtual machines, running multi-CPU architectures. Instead of
    running a local R script with the local CPU and RAM restrictions,
    the same code can be processed on all CPU threads of the cluster of
    machines in the cloud, all running in a Docker container with the
    same R environments. This is achieved through
    \texttt{googleComputeEngineR}'s integration with the R parralisation
    library
    \href{https://CRAN.R-project.org/package=future}{\texttt{library(future)}
    by Henrik Bengtsson}. Local R computation can be thrown up to a
    multi-CPU and VM environment to achieve parrell computation in a few
    lines of R code in your local session. -
    \href{https://cloudyr.github.io/googleComputeEngineR/articles/massive-parallel.html}{some
    demonstrations are available here}.
  \end{itemize}
\item
  \href{https://cloud.run}{Google Cloud Run} - This is a CaaS
  (Containers as a Service) that lets you launch a Docker container
  without worrying about underlying infrastructure. This dispenses with
  the developer creating a cloud server to run the Docker image on, by
  abstracting away those servers to a more serverless configuration.
  Cloud Run lets you run your code on top of a managed or your own
  Kubernetes cluster running Knative, and can accept any Docker image.
  The service takes care of network ingress, scaling machines up and
  down to zero, authentication and authorisation, all features which are
  non-trivial for a developer to create on their own. This can be used
  to scale up R code to millions of instances if they need to, with
  little or no changes to existing code. An R implementation is shown
  here at
  \href{https://github.com/MarkEdmondson1234/cloudRunR}{cloudRunR} which
  uses Cloud Run to create a scalable R plumber API.

  \begin{itemize}
  \tightlist
  \item
    How does this relate to
    \url{https://code.markedmondson.me/googleCloudRunner/index.html} ?`
  \end{itemize}
\item
  \texttt{batchtools} \citep{Lang2017batchtools} can
  \href{https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html}{schedule
  jobs with Docker Swarm}
\item
  scalable deployments, e.g., start with numerous Shiny talks mentioning
  Rocker at useR!2017
\item
  \href{https://github.com/dynverse/dynmethods}{dynmethods}
  \citep{rcannood}: In order to evaluate ±50 computational methods which
  all used different environments (R, Python, C++, \ldots{}), we wrapped
  each of them in a docker container and can execute these methods from
  R. Again, all of these containers are being built on travis-ci, and
  will only be pushed to docker hub if the integration test succeeds.
\item
  \pkg{drake},
  \url{https://docs.ropensci.org/drake/index.html?q=docker\#with-docker}
\end{itemize}

\hypertarget{packaging-programs-for-processing-pipelines}{%
\subsection{Packaging programs for processing
pipelines}\label{packaging-programs-for-processing-pipelines}}

\label{packaging}

Docker is perfectly suited to package and execute any software and data,
and is a good candidate to build complex processing pipelines
independent of the used programming language. Because of the original
use case (see~\nameref{introduction}), Docker does not have a standard
mechanism to chain a number of containers together, i.e., to define how
environment variables, volume mounts, or ports can be used to pass input
parameters and data into a container and how to get results out. Some
packages, e.g.~\pkg{containerit}, provide a Docker image that can be
used similar as a CLI, but the usage is
cumbersome\footnote{\href{https://o2r.info/containerit/articles/container.html}{https://o2r.info/containerit/articles/container.html}}.

\textbf{\pkg{outsider}} (\url{https://antonellilab.github.io/outsider/})
tackles the problem of integrating external programmes into an R
workflow \citep{bennett_outsider_2019}. Installation and usage of
external programmes can be cumbersome and even prohibited if platform
dependent. Therefore \pkg{outsider} uses the platform-independent Docker
images to encapsulate processes in \emph{outsider modules}. Each
outsider modules has a \texttt{Dockerfile} and an R package with
functions for interacting with the encapsulated tool. A user only needs
to use the given functions to install a module and can then integrate a
tool into their own workflow seaminglessly using only R functions. The
module takes care of transmitting arguments and sending and receiving
files to and from a container. Outsider modules are hosted, e.g., on
GitHub, and \pkg{outsider} contains discovery functions for them.

\hypertarget{research-compendia}{%
\subsection{Research Compendia}\label{research-compendia}}

\label{compendia}

\begin{itemize}
\tightlist
\item
  researchcompendia.science
\item
  \url{https://github.com/benmarwick/rrtools}
\item
  \pkg{renv}, \url{https://rstudio.github.io/renv/articles/docker.html}
\item
  what to copy in and what not
\end{itemize}

@@

Containers can also serve as useful playgrounds to create environments
ad-hoc or to provide very specific environments that are not needed or
not easily available in day-to-day development. These environments may
have specific versions of R, of R extension packages, and of system
libraries used by R extension packages, and all of the above in a
specific combination.

First, such containers can greatly facilitate \textbf{fixing bugs},
because developers can readily start and later discard a container to
investigate a bug report without affecting their regular system, and
potentially having to revert changes back. Using the Rocker images with
RStudio, these disposable environments lack no development comfort
(cf.~Section~\nameref{compendia}). \citet{eddelbuettel_debugging_2019}
describes an example how a Docker container was used to debug an issue
with a package only occuring with a particular version of Fortran, and
using tools which are not readily available on all platforms (e.g., not
on macOS). For the \href{https://www.r-spatial.org/}{R-spatial
community}, using containers is particularly useful, because of the
strong integration of system libraries in core packages for geospatial
data modelling and analysis. Purpose built Docker are used to prepare
for upcoming releases of system libraries, individual bug reports, and
for the lowest supported versions of system libraries
\textbackslash{}footnote\{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker},
\textbackslash{}href\{\url{https://github.com/Nowosad/rspatial_proj6}\}\{\url{https://github.com/Nowosad/rspatial_proj6}\},
\href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}\}.

There are also special images for identifying problems beyond the mere R
code, such as \textbf{debugging R memory problems}. The images
significantly reduce the barrier to follow complex steps for fixing
memory allocation bugs \citep[cf. Section~4.3 in][]{core_writing_1999}.
These problems are hard to debug and critical, both because when they
occur they lead to fatal crashing processes.
\href{https://github.com/rocker-org/r-devel-san}{\texttt{rocker/r-devel-san}}
and
\href{https://github.com/rocker-org/r-devel-san-clang}{\texttt{rocker/r-devel-ubsan-clang}}
are Docker images have a particularly configured version of R to trace
such problems with gcc and clang compilers, respectively
\citep[cf.~\pkg{sanitizers} for examples,][]{eddelbuettel_sanitizers_2014}.
The image \href{https://github.com/wch/r-debug}{\texttt{wch/r-debug}} is
a purpose built Docker image with \emph{multiple} instrumented builds of
R, each with a different diagnostic utility activated.

Second, containers are useful for \textbf{testing} R code during
development. To submit a package to CRAN, an R package must work with
the development version of R, which must be compiled locally. That can
be a challenge for some users. The R-hub service (see
Section~\nameref{rhub}) makes it easy to ensure that no errors occur,
but to fix errors a local setup is still often warranted, e.g.~using the
image \texttt{rocker/r-devel}, and to test packages with native code,
which can make the process more complex
\citep[cf.][]{eckert_building_2018}. The R-hub Docker images can also be
used to debug problems locally using various combinations of Linux
platforms, R versions, and
compilers\footnote{See \href{https://r-hub.github.io/rhub/articles/local-debugging.html}{https://r-hub.github.io/rhub/articles/local-debugging.html} and \href{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}{https://blog.r-hub.io/2019/04/25/r-devel-linux-x86-64-debian-clang/}}.
The images go beyond the configurations, or \emph{flavours}, used by
CRAN for checking
packages\footnote{\href{https://cran.r-project.org/web/checks/check_flavors.html}{https://cran.r-project.org/web/checks/check\_flavors.html}},
e.g., with CentOS-based images, but lack a container for checking on
Windows or OS X. The images greatly support package developers to
provide support on operating systems they are not familiar. The package
\pkg{dockertest} (\url{https://github.com/traitecoevo/dockertest/}) is a
proof of concept for automatically generating \texttt{Dockerfile}s and
building images specifically to run
tests\footnote{\pkg{dockertest} is not actively maintained, but mentioned still because of its interesting approach.}.
These images are accompanied with a special launch script so the tested
source code is not stored in the image but the currently checked in
version from a local Git repository is cloned into the container at
runtime. This approach clearly seperates test environment, test code,
and current working copy of the code.
